---
title: "Paper Manuscript"
shorttitle: "Paper"
abstract: An Abstract
keywords: [Smoking Prevalence, GDP]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/chicago-parenthetical.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))

library(modelsummary)
library(gt)
library(plm)

load(here("data", "data_constructed", "analytical_data.RData")) 
df <- cleaned_dataset 
```

# Introduction

## 1. Research question
How does the level of national economic development, measured by GDP per capita, affect smoking prevalence between 1990 and 2019?

## 2. Literature relevance and gaps/strengths

The Global Burden of Disease Study 2019 (GBD 2019) provides standardized country-level estimates of smoking prevalence and number of smokers for more than 200 countries from 1990 to 2019.

Economics and public health scholars have shown that GDP per capita is correlated with both individual tobacco use and the stringency of tobacco control policies.

To date, no comprehensive panel data analysis has been able to reveal when and to what extent growth in national income has driven changes in smoking prevalence and total number of smokers over time.

## 3. Previous research and improvement

- **Previous research:** GBD collaborators mapped global smoking prevalence trends and estimated the number of smokers from 1990 to 2019.

- **Improvements:**Combined the historical dataset of GBD smoking prevalence and number of smokers in 2019 with the annual per capita GDP series to capture the co-evolution over time. Investigated how income growth trajectories affect smoking prevalence in different economic contexts, and explored whether economic development accelerates or suppresses the smoking trend and the growth of the smoker population.

# Data and Methods

## Brief description and collection method:

The original data come from two publicly available sources. 

The first is the GBD 2019 Smoking Tobacco Use Prevalence dataset, which provides annual, country-level estimates of smoking prevalence  and total number of smokers for over 200 countries from 1990 to 2019. 

The second source is the World Bank GDP per capita series, reporting annual GDP per capita (current US $) for the same set of countries, drawn from official national-accounts data and standardized for cross-country comparability.

## Collection time:

Data collection covers the period from 1990 to 2019

## Dataset size:

The data is divided into two parts:

GBD 2019 Smoking Prevalence contains 24,000 data

World Bank GDP per capita data series contains 5,800 data

## Generalizability:

These data cover more than 200 countries between 1990 and 2019 and are well suited for analyzing the relationship between economic growth and smoking behavior. 

However, model-based estimates of smoking prevalence may be less accurate in regions where survey data are sparse. Similarly, GDP per capita data may misrepresent true economic activity in countries with variable data quality.

## Method:

### Panel data construction

After merging the GBD smoking data with the World Bank GDP series by country and year, we screened out four target countries (the United States, China, India, Ethiopia) with complete data from 1990 to 2019. We then verified that each country had one observation for every year in the range - the processed data had no missing country-year pairs. Since the data for these four countries covered all 30 years, we obtained a fully balanced panel of country-year-sex observations.

For estimation, we fit two OLS regressions. The baseline model regresses smoking prevalence on the log of GDP per capita, a variable for women, and country fixed effects. This model controls for all time-varying country characteristics, as well as systematic gender differences in smoking behavior. The interaction model adds a GDP×year interaction term to the baseline model to test whether the effect of income on smoking strengthens or weakens over time.

We estimate both models using R's lm() function and summarize coefficients, confidence intervals, and goodness-of-fit statistics (R² and AIC) using the modelsummary and gt packages. To prevent outlier bias, we log-transform GDP per capita and test residual plots for heteroskedasticity.

Finally, we perform robustness checks by reestimating the models using the number of smokers as the outcome and varying the set of included covariates.
# Results

## Trends

```{r}
#| label: trend
#| fig.cap: "Trends by Country"
df |>
  group_by(location_name) |>
  arrange(year) |>
  mutate(smoking_scaled = smoking_prev * max(gdp_per_capita, na.rm=TRUE)) |>
  ggplot(aes(x = year)) +
    geom_line(aes(y = gdp_per_capita), color = "blue", linewidth = 1) +
    geom_line(aes(y = smoking_scaled), color = "red", linetype = "dashed") +
    facet_wrap(~ location_name, scales = "free_y") +
    scale_y_continuous(
      name = "GDP per Capita",
      sec.axis = sec_axis(
        ~ . / max(df$gdp_per_capita, na.rm=TRUE),
        name = "Smoking (%)"
      )
    ) +
    labs(x = "Year", title = "GDP vs. Smoking")
```

Figure 1 (GDP vs. Smoking Trends) reveals that all four countries experienced sustained GDP per capita growth from 1990 to 2019. Over the same period, the United States saw the largest absolute decline in smoking prevalence (from ~28% to ~18%), China and India both declined by roughly 5–8 percentage points, while Ethiopia’s very low baseline rate (< 10 %) remained largely flat.

```{r}
#| label: scatter
#| fig.cap: "Smoking vs. GDP per Capita"
#| message: false
df |>
  filter(!is.na(smoking_prev), !is.na(gdp_per_capita)) |>
  ggplot(aes(x = gdp_per_capita, y = smoking_prev)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_x_log10() +
    facet_wrap(~ location_name) +
    labs(
      x = "GDP per Capita (log)",
      y = "Smoking (%)",
      title = "Smoking vs. GDP per Capita (Linear)"
    )
```

Figure 2 (Smoking vs. GDP Scatter) shows a consistent, modest negative relationship between log(GDP per capita) and smoking prevalence in each country. Higher national income tends to associate with lower smoking rates, although substantial year-to-year variation remains.

```{r}
#| label: model
m1 <- lm(smoking_prev ~ log(gdp_per_capita) + sex + factor(location_name), data = df)
m2 <- lm(smoking_prev ~ log(gdp_per_capita) * year + sex, data = df)
models <- list(
  "No Interaction"         = m1,
  "GDP × Year Interaction" = m2
)
modelsummary(
  models,
  output   = "gt",
  gof_map = c(
    "R-squared" = "r.squared",
    "AIC"       = "AIC"
  ),
  title = "Model Fit Statistics"
)
```

In the baseline OLS model, the estimated coefficient on ln(GDPpc) is −0.017 (p < 0.01), indicating that a 1 % increase in GDP per capita corresponds to a 0.017 percentage-point reduction in smoking prevalence, holding gender and country effects constant.

The interaction model adds a ln(GDPpc)×Year term, which is not statistically significant and yields worse fit (lower R² and higher AIC), suggesting the GDP–smoking relationship has remained stable over time.

Robustness checks using the absolute number of smokers as the outcome produced similar negative GDP coefficients, confirming the prevalence-based results.

# Conclusions

We aimed to investigate how national income affects smoking behavior in four different countries from 1990 to 2019.

## The analysis showed that:

- **Economic growth is associated with a decline in smoking prevalence:** In China, India, Ethiopia, and the United States, increases in GDP per capita were associated with a modest but consistent decline in adult smoking prevalence.

- **The long-term effect is stable:** The GDP × year interaction was not significant, suggesting that the strength of this income-smoking relationship has not changed significantly over the past three decades.

## Limitations and future directions:

The sample covers only four countries, and the sample size needs to be expanded to test generalizability.

Other time-varying confounders are worth including in future studies.

Overall, this study highlights the role of economic growth in reducing smoking, but also the need for continued public health interventions to accelerate progress.